---
title: "Introduction to epitable"
#author: "Vignette Author"
#date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: 
      - !expr system.file("extdata", "epi-chart-vignette.css" ,package = "epitable")
vignette: >
  %\VignetteIndexEntry{Introduction to epitable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)

opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})


```
epitable will print HTML tables according to EPI website specifications.

To install, 
```{r eval=FALSE}
devtools::install_github("Economic/epitable")

## install package and vignettes
devtools::install_github("Economic/epitable", build_opts = c("--no-resave-data", "--no-manual"))
browseVignettes("epitable")
```

## Basic example
epitable contains some example data on trade balances:
```{r}
library(epitable)
tradebalance
```

Here is the simplest use of epitable:
```{r}
epitable(tradebalance, rownamesvar = industry)
```

Note that you must specify the row names, either as a variable in the above, or as vector:
```{r, message=FALSE}
library(dplyr)
myrownames <- paste(c("My row"),seq(1:12))
tradebalance %>% select(-industry) %>% epitable(rownames = myrownames)
```

## Output options
As you've noticed, by default epitable's output shows up in the RStudio viewer. You can also write the table snippet to a file:
```{r eval=FALSE}
epitable(tradebalance, rownamesvar=industry), file = "mytable.html")
```

That file will only contain the snippet of HTML that is the table. To write a fully self-contained web page, use the selfcontained option:
```{r eval=FALSE}
epitable(tradebalance, rownamesvar=industry), file = "mytable.html", selfcontained=TRUE)
```

To display the raw HTML of the table snippet in the console, try

```{r test, output.lines=10}
print(epitable(tradebalance, rownamesvar=industry), useViewer=FALSE)
```

## Formatting entries (use the scales package)
An easy way to format decimal points, percent signs, etc., is to use the scales package, which transforms columns of your dataset into character strings:

```{r}
library(scales)

tradebalance_fmt <- tradebalance %>%  
  mutate_at(vars(ends_with("level")), ~number(.,accuracy=0.1, big.mark=",")) %>% 
  mutate_at(vars(ends_with("share")), ~percent(.,accuracy=0.1))

epitable(tradebalance_fmt, rownamesvar=industry)
```

## Rows
Row name indendation is handled by the rowlevels option:
```{r}
epitable(tradebalance_fmt, rownamesvar=industry, rowlevels=rep(1:4,3))
```

Here is an example of how you might do this a little less randomly:
```{r}
thelevels <- tradebalance_fmt %>%
  mutate(level = 3) %>%
  mutate(level = ifelse(grepl("Total change|Agriculture, forestry|Mining|Manufacturing", .$industry), 1, level)) %>%
  mutate(level = ifelse(grepl("Oil and gas|Minerals and ores|Nondurable goods|Industrial supplies", .$industry), 2, level)) %>% 
  select(level) %>% 
  pull()

epitable(tradebalance_fmt, rownamesvar=industry, rowlevels=thelevels)
```

## Columns
``` {r}
epitable(tradebalance_fmt, rownamesvar=industry, rowlevels=thelevels, colnames=c("Import levels", "Import shares", "Export levels", "Export shares", "Trade balance levels", "Trade balance shares"))
```

``` {r}
epitable(tradebalance_fmt, rownamesvar=industry, rowlevels=thelevels, colgroups=c("Imports", "Exports", "Trade balance"), colgroupspattern=c(3, 1, 2), colnames=rep(c("Levels (\\$billions)", "Share of total change"),3))
```

## More Examples

```{r, echo=TRUE, results='asis'}
library(dplyr)
library(scales)
thelevels <- tradebalance %>%
  mutate(level = 3) %>%
  mutate(level = ifelse(grepl("Total change|Agriculture, forestry|Mining|Manufacturing", .$industry), 1, level)) %>%
  mutate(level = ifelse(grepl("Oil and gas|Minerals and ores|Nondurable goods|Industrial supplies", .$industry), 2, level))

tradebalance_fmt <- tradebalance %>%  
  mutate_at(vars(ends_with("level")), ~number(.,accuracy=0.1, big.mark=",")) %>% 
  mutate_at(vars(ends_with("share")), ~percent(.,accuracy=0.1))

epitable(tradebalance_fmt, rownamesvar=industry, rowlevels=thelevels$level, colgroups=c("span 1", "span 2", "span 3"), colgroupspattern=c(3, 1, 2), colnames=c("import levels", "import shares", "import levels", "import shares", "import levels", "import shares"))
```

